---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TPSClus

<!-- badges: start -->
<!-- badges: end -->

`TPSClus` is an R package designed to implement fuzzy clustering of multivariate longitudinal data using tensor product smoothing splines. The `TPSfit()` function fits the splines to three or more variables that are measured repeatedly over time. The `cluster.fitted()` or `cluster.coefs()` functions can then be used to identify fuzzy clusters in the data. Lastly, the `FKM.glm()` function can be used to generate a generalized linear model for a distal outcome using the fuzzy clusters and other covariates as predictors. A number of plot options are available for visualization of the data and clusters.

## Installation

You can install the development version of TPSClus like so (need to update):

``` r
devtools::install_github("silivingston/TPSClus", dependencies = TRUE, build_vignettes = TRUE)
```

## Example

Basic example of flow using built-in simulated dataset:

Fit splines to multivariate data:
```{r splines}
library(TPSClus)
data(TS.sim)

fitsplines2 <- TPSfit(TS.sim, vars=c("Var1", "Var2", "Var3"),
time="Time", ID="SubjectID", knots_time=c(0, 91, 182, 273, 365),
     fit_times=c(46, 91, 137, 182, 228, 273, 319))
```

Run fuzzy clustering algorithm on the smoothed data using 3 clusters:
```{r clustering}
clusters1 <- cluster.fitted(fitsplines2, k=3, m=1.3, seed=12345, RS=5, noise=TRUE)
summary(clusters1)
```

Plot the mean cluster trajectories:
```{r plot}
plot(clusters1, legend_label_size=10)
```

Fit a GLM using the clusters and other covariates as predictors:
```{r model}
model <- FKM.glm(clusters1, TS.sim, y="outcome", covariates=c("x1", "x2"),
family="binomial")
summary(model)
```

Predict the outcome for a new dataset:
```{r predict}
data(TS.sim.new)
predicted <- predict(model, TS.sim.new)
head(predicted)
```

<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>. -->

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo = FALSE} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN. -->
